### 2018-02-03 thesis/contents/tn_sagit.tex

\textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with up to 80\% accuracy from controls. Affected CN V differences were more focused in the cistern and root-transition-zone, whereas unaffected CN V show more focus towards the brainstem. The affected S1 showed focused MD differences towards the cortex with an accuracy of 85\%.

\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN vs. control and TN white-matter machine learning classification using GP. The method reproduced previous findings with finer-grained regional specificity. The study also revealed specific diffusivity changes in deeper white matter.

\section{Introduction}
\subsubsection{Trigeminal Neuralgia}
Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches. TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}. The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}. We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is disrupted in MS-TN. In all previous TN studies, placements of region-of-interests in the CN V were pre-determined and manually placed. In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.

The trigeminal sensory system involves both the discriminative touch and pain pathways. Anatomicall, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex. The pain pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord. The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus. The pain pathway also projects to the medial nucleus of the thalamus, where it then project medially to the cingulate cortex.

\subsubsection{Data Analysis from Streamlines}
White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. Obtaining diffusivity metrics from tractography involves converting streamline bundles into a volumetric spatial mask, followed by measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The naive form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-through. This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm. In white matter structures such as CN V as well as other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

@@

The principal advantage of a streamline-based approach is that we can retain the pathway orientation from its geometry. White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning. We then merge the tracts into a template space using non-linear deformations, and  perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.

\subsubsection{Gaussian Process}
The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}. However, one disadvantage of these networks is in its inability to quantify uncertainty and provide interpretable models. Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}. In fact, it has been found that DNN with an infinite number of hidden units approximates GP \cite{neal1996priors}.

@@

\section{Methods}
\subsubsection{Acquisitions}
We studied 36 patients with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).  All TN patients were scanned at 1 day pre-GK and 6 months post-GK. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256. DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}. We deployed selective automatic group integrated tractography (SAGIT) framework to automate preprocessing, registration, and tract generation. The processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

\subsubsection{Tractography}
ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI: cistern nerve root; excludes: cerebellum grey matter, pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs: trigeminal nucleus (TGN); includes: contralateral VC thalamus, pontine decussation; excludes: ipsilateral thalamus, cerebellar grey matter, pons, contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs: VC thalamus; includes: ipsilateral S1 white-matter boundary; excludes: brainstem).

Constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} was used to delineate the three sets of anatomy. Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy. Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration. Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = 45$\deg$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum lengh = 80 mm, initial cut-off = 0.15, cut-off = 0.15. Probabilistic tractography for TPT project was performed with step size = 1mm, and streamline count = 2000, and maximal length = 60 mm.

Tractography was first performed in the native DWI space for each subject. Diffusivity metrics of FA, AD, RD, MD, and GFA  were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation. Additionally, subject ID and group ID scalar values were also inserted into each tractography files for identification.

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. A software library called “Fascicles” is created by the author to facilitate this process. The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported into a comma-delimited file. The points were then transformed by applying the subject DWI to T1, and T1 to template transforms using ANTs’  applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the SQLite database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved. Finally, all template-space tractography models across all the subjects are imported into a new SQLite database, and then exported as a single file for analysis.

\subsubsection{Along-the-tract Analysis}
The merged tract was first post-processed in 3D Slicer to remove erroneous and stray streamlines that will confound measurement. Streamlines removed include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise. S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed. The TPT streamlines were not examined due to inconsistent reproducibility across subjects.

Tractography centroids, which were the average curve best fit the tract bundle, was defined on the merged tracts in template space. CN V centroids were manually defined, using 3 control points situated in the template world space. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm. The spline is then resampled into 30 points that become the centroid definition with segment length of 1.3 mm. Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids. Diffusion metrics from the points were aggregated for each of the cross-sections.  S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, with centroid length of 62.1 mm and segment length of 2.07 mm.

@@

Diffusivity metrics of affected (right side) and unaffected CN V and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides. The affected and unaffected sides in TN patients were also compared with each other. 
Age was regressed out by using the residuals of the linear regression model from the controls group. The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training. 
Gaussian Process (GP) classification was exhaustively performed on CN V and S1 nerve segments to identify key segment differences. Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment. The entire machine learning process was performed using Scikit-learn machine learning software framework.
For each model, the subjects were split into train and validation sets using stratified 10-fold cross-validation. The GP model was initiated with Matern kernel (length scale = 1), and the model fitted using the training set. The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed segment. Segments accuracy greater than 75\% were accepted, and the segment length with the greatest accuracy was then determined.

\subsection{Results}
\subsubsection{Group Tractography}
Merged CN V delineations show anatomical consistency, and extended into the ipsilateral TGN. Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration ( Figure \ref{fig:GPfigure1}). 
The TPT projections, when filtered strictly from the TGN to the ipsilateral VC thalamus region, showed distinct bundles of decussating pathways between left and right. The resulting pathway was not symmetrical, where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}). TPT streams lines were sparse, suggest low inter-subject reproducibility. Therefore no additional analysis was carried out for TPT.
S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}).

@@

\subsubsection{CN V}
Both sides were distinguished from controls with classifiers that achieve greater than 75\% accuracy. 
Classifiers on the right side have performance ranging from 77\% (GFA) to 80\% (MD) accuracy. The left classifiers achieved accuracy ranging from 75\% (FA) to 80\%(GFA). The right side shows more consistent overlap of segment ranges between positions 10--17, in the region of the cistern/REZ; while the left side shows less consistent range overlaps, and AD segment range from positions 3 to 14, which include the brainstem regions. Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extends more into the brainstem. Comparing left/right, only the FA classifier achieved greater than 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment (Figure \ref{fig:GPfigure4}).

The best performing segments versus controls is GFA (segment 8--14) on the left (unaffected) side, with 80\% accuracy, 84\% precision, and AUC of 0.8; while MD (segment 10--17) performed the best on the right (affected) side, with 80\% accuracy, 84\% precision, and AUC of 0.79 (Table \ref{table:CN V}, Figure \ref{fig:GPfigure6}). The best performing left-versus-right classifier was FA (segment 11--19), with 75\% accuracy, 82\% precision, and AUC of 0.75.

@@

\subsubsection{S1}
Both sides of the S1 face region pathway were able to be distinguished by classifiers with greater than 75\% accuracy. 
The FA and RD classifiers at locations 617, representing the segment that emerge from the thalamus and merging into corona radiata, are common to both sides. The left side has higher classification accuracy of 80\% with FA, while the right side had a maximum accuracy of about 78\% with RD. 
The MD classifier meanwhile  is unique to the left side, and achieved an accuracy of 85\% in locations 16--23, near the grey-white matter boundary, closer to the cortex (Figure \ref{fig:GPfigure5}).

The best performing segment versus controls was MD (segment 16--23) on the left (affected) side, with 85\% accuracy, 88\% precision, and AUC of 0.83; whereas the best on the right (unaffected) side was FA (segment 6--16), with 78\% accuracy, 81\% precision, and AUC of 0.76 (Table \ref{table:s1}, Figure \ref{fig:GPfigure6}).

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{figure-roc.pdf}
\caption{ROC curves of the top performing classifiers}
\label{fig:GPfigure6}
\end{figure}

>>>>

\textbf{Results:} Both affected and unaffected trigeminal nerves in TN patients can be differentiated with 80\% accuracy from controls. Affected CN V FA segments that are differentiated from trons have more overlap towards the distal segments in the cistern and gangion, whereas unaffected CN V segment overlaps extend more towards the brainstem. The bilateral middle segment of S1 can be differentiated from controls. Where the affected S1 RD classifier achieved higher accuracy of 87\%.

\textbf{Conclusions:} We  demonstrate the first instance of group-wise merged tractography analysis of TN vs. control and TN white-matter machine learning classification using GP. The method reproduced previous findings with finer-grained regional specificity. The study also revealed specific diffusivity changes in TPT and S1 white matter pathways.

\section{Introduction}
\subsubsection{Trigeminal Neuralgia}
Trigeminal neuralgia (TN) is a debilitating facial neuropathic pain syndrome characterized by paroxysmal and shock-like pain in one or more dermatomes of the three trigeminal nerve (The fifth cranial nerve: CN V) branches. TN manifests most commonly as classic (also called idiopathic TN), or secondary to a range of neuropathology, including tumors such as meningiomas \cite{Cheng2008}, trigeminal schwannomas \cite{Miller2008}, neurocysticercosis \cite{Revuelta1995}, and most notably multiple-sclerosis (MS) \cite{Cruccu2009,VanderMeijs2002,Nick2012}. The pathophysiology of idiopathic TN is thought to be vascular compression at the root of the CN V nerve entry zone (NEZ) \cite{Linn2011,Love2001}, although idiopathic TN can also occur without evidence of vascular compression \cite{Lee2014}. We previously compared TN and MS-TN diffusivity differences in four segments of the CN V at the level of the pons \cite{Chen2016a}, and confirmed that brainstem CN V diffusivity is altered in MS-TN.

The trigeminal sensory system involves both the discriminative touch and pain pathways. Anatomically, the discriminative pathway synapses onto the primary trigeminal sensory nucleus, and then projects to the contralateral VC thalamus and then to the S1 cortex. The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord. The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus. The pain pathway also projects to the medial nucleus of the thalamus, where it then project medially to the cingulate cortex.

\subsubsection{Along-the-tract analysis}
White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. Obtaining diffusivity metrics from tractography involves converting streamline bundles into a volumetric spatial mask, followed by measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The simplest form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-throughs. This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm. In white matter structures such as CN V as well as other small pathways, the structure diameter may be no more than 3mm, and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

@@

The principal advantage of a streamline-based approach is that we can preserve the fiber orientations. White matter metrics in 3D space can then be projected to a lower dimensional centroid for detailed analysis. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework (Chen et al., 2016) to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning. We then merge the tracts into a template space using non-linear deformations, and  perform tractography centroid identification and statistics at the merged group level to ensure anatomical consistency.

\subsubsection{Gaussian Process}
The breakthrough results of deep convolutional neural networks in computer vision \cite{Krizhevsky2012} have sparked the interest of deploying similar systems for medical imaging diagnosis \cite{Greenspan2016}. However, one disadvantage of deep neural networks is in its inability to quantify uncertainty and provide interpretable models. Model interpretability is a fundamental requirement for machine learning in medicine, and therefore there is renewed interest in bayesian-based models like Gaussian Process (GP) \cite{gal2016dropout}.

@@

In all previous TN studies, placements of measurement region-of-interests in the CN V were pre-planned and manually placed by researchers. Manual tractography generation with manual ROI placements was a roadblock in performing more indepth group-wise tractography examination of the trigeminal pathways. We had developed and validated the SAGIT tractography framework to allow automated group tractography \cite{Chen2016}. In this study we further expand the study to A) analyze the CN V differences between TN and healthy controls across the entire nerve using along-the-track and machine-learning methods to auto-discover key regions-of-interest; B) extend the examined regions to include trigeminopontothalamic (TPT) and thalamocortical (S1) white matter pathways as part of the trigeminal sensory pathways.

\section{Methods}
\subsubsection{Acquisitions}
We studied 36 patients with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).  All patients underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256. DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains \cite{Chen2016}. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}. We deployed selective automatic group integrated tractography (SAGIT) framework \cite{Chen2016} to automate preprocessing, registration, and tract generation. The SAGIT processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

\subsubsection{Tractography}
ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibers must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary; it must exclude the brainstem).

We used constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} to delineate the three defined anatomical regions. Deterministic and probabilistic streamline algorithms were used to delineate different regions, based on findings in the SAGIT study \cite{Chen2016}. Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy. The parameters for each were as follows: Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration. Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = 45$\deg$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum lengh = 80 mm, initial cut-off = 0.15, cut-off = 0.15. Probabilistic tractography for TPT project was performed with step size = 1mm, and streamline count = 2000, and maximal length = 60 mm. Tractography was first performed in the native DWI space for each subject.  FA, AD, RD, MD, and GFA were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. A software library\footnote{https://github.com/sinkpoint/fascicle} is created by the author to facilitate this process, the steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group ID associations preserved. Finally, all template-space tractography models across all the subjects are merged into a single file.

\subsubsection{Along-the-tract Analysis}
The merged tract was post-processed in 3D Slicer to remove erroneous streamlines that confounded measurement. Streamlines removed include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise. TPT streamlines that did not decussate at the level of the pons, as well as fibers that stray into the contralateral TGN were removed. S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed.

Tractography centroids, which were the average curve best fit the tract bundle, were identified on the merged tracts in template space. CN V centroids were manually defined using 3 control points. A Catmull-Rom spline \cite{DeRose1988} was then defined from these control points using the VTK software library \cite{Schroeder2005}, which resulted in a curve of length 38.91 mm. The spline is then resampled into 30 points that become the centroid definition with segment length of 1.3 mm. Nearest-neighbor point assignment was used to divide the merged tract into 30 cross sections along the centroids. Diffusion metrics from the points were aggregated for each of the cross-sections.  S1 centroids were defined using QuickBundles \cite{Garyfallidis2012} with distance threshold of 30, resulting in centroid length of 62.1 mm and segment length of 2.07 mm.

@@

Diffusivity metrics of affected (right side) and unaffected CN V and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides. The affected and unaffected sides in TN patients were also compared with each other. 
Age was regressed out by using the residuals of the linear regression model from the controls group. The diffusion values were subsequently normalized to zero mean with ranges $[-1, 1]$ for model training. 
Gaussian Process (GP) classification was performed on CN V and S1 nerve segments to identify key segment differences. Multiple moving windows of length between 5 to 30 were used to constraint tract segments, and a Gaussian Process model was trained for each windowed segment. The entire machine learning process was performed using Scikit-learn machine learning software framework.
For each model, the subjects were split into train and validation sets using stratified 10-fold cross-validation. The GP model was initiated with Matern kernel (length scale = 1), and the model fitted using the training set. The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed segment. Segments accuracy greater than 75\% were accepted, and the segment length with the greatest accuracy was then determined.

\subsection{Results}
\subsubsection{Tractography}
Merged CN V delineations show anatomical consistency, and extended into the ipsilateral TGN. Changes in FA along the streamlines demonstrate clear differences between cistern and brainstem segments, suggesting good agreement in anatomical registration ( Figure \ref{fig:GPfigure1}). 
The TPT projections, when filtered strictly from the TGN to the ipsilateral VC thalamus region, showed distinct bundles of decussating pathways between left and right. The resulting pathway was not symmetrical, where the right TPT pathway diverts and envelopes the left (Figure \ref{fig:GPfigure2}).
S1 streamlines reliably extended into the face region of the S1 cortex (Figure \ref{fig:GPfigure3}).

@@

\subsubsection{Diffusivity Differences}
\paragraph{CN V}
Both sides were distinguished from controls with classifiers that achieve around 80\% accuracy (Figure \ref{fig:GPfigure4}). 
Classifiers on the right side have performance ranging from 76\% (GFA) to 86\% (FA) accuracy. The left classifiers achieved accuracy ranging from 79\% (MD) to 81\%(FA). The right side shows more overlap of segment ranges between positions 14--24, in the region of the cistern/gangion; while the left side shows segments that covers the entire the nerve, and extends to cover position 0--4 in the brainstem. Overall, the right side features extend further into the distal space, into the cistern and even the ganglion, while the left side extends more into the brainstem. FA consistently achieved the highest accuracy in both sides, with right FA at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cister).
Comparing left/right, the FA classifier achieved 75\% accuracy, where its segments range from 11 to 19, which is the cisternal segment. The GFA classifier achieved 70\% accuracy in region 9--12. Demonstrating that affected-unaffected diffusivity differences are found in the cistern/REZ (Table \ref{table:CN V}).

@@

\paragraph{Trigeminopontothalamic tract}
Both left and right TPT pathways can be distinguished from controls with greater than 80\% accuracy (Figure \ref{fig:GPfigureTPT}). The right maximum accuracy was 85\% (FA), and minimum accuracy 83\% (RD), while the left maximum accuracy was 89\% (FA) and minimum 71\% (RD). The regions covered both the decussation and the ascending segments of the TPT, with the right decussation showing more proximal coverage. The left-right classifiers achieved maximum accuracy of 98\% (FA) and minimum of 88\% (MD). The left-right FA covered position 4--14, suggesting strong laterality differences in the TPT (Table \ref{table:tpt}).

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{figure-GP-TPT.pdf}
\caption{TPT classifier performance}
\label{fig:GPfigureTPT}
\end{figure}

\begin{table}[ht]
\centering
\csvautotabular{tpt_table.csv}
\caption{TPT GP classifiers performance data}
\caption*{List of the best accuracies for each diffusion metric. Precision, recall, and f1 scores are also provided for reference}
\label{table:tpt}
\end{table}

\paragraph{S1}
Both sides of the S1 face region pathway were able to be distinguished by classifiers with about 85\% accuracy. The left RD classifier achieved max accuracy of 87\% at position 13--20, while the right RD classifier achieved max accuracy of 83\% at psotion 8--26 (Figure \ref{fig:GPfigure5}). Overall the left classifier ranged from 72\% (AD) to 87\% (RD), and the right classifieres ranged from 73\% (GFA) to 83\% (RD) (Table \ref{table:s1}). Both sides differ from controls in the branching region between the thalamus and cortex. Left-right classifiers showed maximum accuracy of 78\% (RD), and minimum of 76\% (FA). The left-right classifiers overlap in the regions 5--10, at the segment where the pathway emerges from the thalamus.

@@
